### 2021-04-09 assertions/doc/thesis-bernhardt/chapters/2-use_case_analysis.tex
IR done; add cost model; cost analysis

\section{Cost analysis}
\draft{Formalize the costs of checking a formula}
\begin{itemize}
\item Describe a general cost model for execution of a SC
\item Calculate costs for checking a formula locally (in a loop)
\item Calculate costs for checking a single element
\item Compare results -> how high is the ``boilerplate'' cost?
\item When do the costs of distributed checking drop below local checking?
\end{itemize}

>>>>

\section{Cost analysis}
In order to be able to evaluate whether distributed assertion checking can indeed lower the cost in comparison to a local verification, the cost of checking a formula on the blockchain needs to be formalized. Since this thesis describes an implementation of the approach on the Tezos blockchain, the following costs refer to that specific gas model, s.t. the results of the cost analyses are comparable. The gas model of Tezos is described briefly in \secref{sec:tezos}. \\
According to the gas model, the total gas consumption of a transaction $T(src,\, dst=C,\, arg, \, ...)$ to an originated contract $C(storage, \, parameter\, type, \, code)$ is calculated by \eqref{eq:gas_transaction}:
\begin{align}\label{eq:gas_transaction}
\begin{split}
GasConsumed_{transaction} = \quad &read(code) \\
+& \, gasRead(storage) \\
+& \, gasDeserialize(storage, \, arg) \\
+& \, gasParseType(parameter \, type, \, storage) \\
+& \, gasParseData(storage, \, arg) \\
+& \, gasParseCode(code) \\
+& \, gasInterpret(code, \, arg, \, storage) \\
+& \, gasUnparse(storage) \\
+& \, gasSerialize(storage) \\
+& \, gasWrite(storage)
\end{split}
\end{align}
Assuming that the storage is never increased, the amount of gas returned by the cost functions independent of the argument, i.e., the parameter value, remain constant for each $T$. The cost functions for the deserialization, parsing and interpretation, however, are functions of $arg$ that grow strictly monotonic with increasing sizes of $arg$. \\
Consider the following simple contract --- it does not do anything useful, but it can be used to approximate the amplification of the cost functions with regard to $arg$. Appendix \ref{apx:cost_analysis_contract} contains the respective Michelson contract. 
\begin{lstlisting}
contract Foo {
	function bar (int a) public {
		for (uint i = 0; i < a; i++) {
			assert(a == a)
		}
		// ...
	}
}
\end{lstlisting}

\draft{Formalize the costs of checking a formula}
\begin{itemize}
\item Describe a general cost model for execution of a SC
\item Calculate costs for checking a formula locally (in a loop)
\item Calculate costs for checking a single element
\item Compare results -> how high is the ``boilerplate'' cost?
\item When do the costs of distributed checking drop below local checking?
\end{itemize}

### 2021-04-09 assertions/doc/thesis-bernhardt/chapters/4-offline_design_tezos.tex
IR done; add cost model; cost analysis

\todo{Add mempool? If Blockchain protocol is outlined}

@@

@@

\subsubsection{High-level languages}
Michelson is a compilation target for various high-level languages that provide a more user-friendly and intuitive way of writing smart contracts. Additionally, some of them come with development environments and testing or verification tools. The following list comprises of the three most prominent languages that compile to Michelson:
\begin{itemize}
\item[\textbf{Liquidity}] \cite{liquidity} is a language with an OCaml-like syntax, which allows using local variables instead of stack manipulations. Its module system can be used to write reusable contract code or libraries. Besides an optimizing compiler, the project also includes a decompiler to compile Michelson programs to Liquidity.
\item[\textbf{SmartPy}] \cite{smartpy} is a language available through a Python library and lets developers write contracts and tests using Python syntax and structures (such as classes). Its developer suite includes i.a. a compiler, a simulation engine for testing contracts and an online editor.
\item [\textbf{fi}] 's \cite{fi} syntax is similar to JavaScript or Ethereum's contract language Solidity. It also provides an online editor and a simulator to test against the compiled contracts. However, it seems extensions to Michelson introduced by newer protocol versions are not supported yet.
\end{itemize}

@@

\subsection{Random}
A vital instruction that cannot be expressed with Michelson's current instruction set (state of protocol 008 Edo \cite{tezos_docs}) is the generation of a random value of certain types. Since smart contracts need to be deterministic, s.t. the network can reach a consensus about the state of the ledger \cite{chatterjee_probabilistic_2019}, the prevalent workarounds for generating pseudo-random numbers cannot be used for assertion checking. Distributed assertion checking explicitly requires as many validators as possible to generate a unique value, whereas common schemes for randomness, like oracles or using block attributes as seeds \cite{chatterjee_probabilistic_2019}, will provide the same values for all validators.\\
\lstref{lst:rand_type} specifies the selection and typing rule of a new instruction \texttt{rand}, which consumes an offset and a positive range from the stack and pushes a randomly generated integer to the top. This instruction, of course, can also be provided for other data types, such as \texttt{nat}, \texttt{string} or \texttt{mutez}.
\lstset{upquote=true}
\begin{lstlisting}[caption=Selection and typing rules of the integer \texttt{rand} instruction, language=, numbers=none, label=lst:rand_type]
:: int : nat: 'A -> int : 'A
> RAND/ offset : range : S  => int : S
\end{lstlisting}

@@

\subsection{Nth}
An exception for not adding higher level instructions may be the more fundamental operation \texttt{nth} for accessing list elements, as it is used (sometimes more than once, ref. \eqref{eq:sorted_v2}) in many use cases. Adding an own predefined primitive for this reduces the origination costs of the assertion contract, given that its equivalent in the current instruction set requires the compiler to generate a workaround using iteration, for instance a loop or list iterator. As a first approximation, appendix \ref{apx:nth} shows a potential implementation of \texttt{nth} in Liquidity (\lstref{lst:nth_liq}) using a loop, and how the respective Michelson code looks like after compiling the contract with the Liquidity compiler (\lstref{lst:nth_tz}). \\
In case the \texttt{nth} is added as a predefined primitive to Michelson for the type \texttt{'a list}, its selection and typing rules are specified as follows:
\begin{lstlisting}[caption=Selection and type rule of the list \texttt{nth} instruction, language=, label=lst:nth_type]
:: (list 'a) : nat : 'A -> option 'a : 'A
> NTH / l : index : S  =>  Some 'a: S
     iff index is within bounds
> NTH / l : index : S  =>  None 'a: S
     iff index is out of  bounds
\end{lstlisting}
As Michelson lists are represented with OCamls list type within the VM \cite{tezos_repo}, the computation for \texttt{nth} could look as shown in \lstref{lst:nth_impl}.
\begin{lstlisting}[caption=Simplified evaluation of \texttt{nth} in the Michelson interpreter, language=, label=lst:nth_impl]
match (instruction, stack) with
...
| (Nth ({elements = []; _}, (_, rest))) ->
  return ((None, rest)
| (Nth ({elements = es; length}, (index, rest))) ->
  let l_length = of_int length in
  if index <= l_length
    then return (Some (List.nth es index), rest)
    else return (None, rest)
\end{lstlisting}
\lstset{upquote=false}

\subsection{failwith 2.0}
\draft{}
- failwith returns a failure string
- assertion execution -> random number is generated
- if counterexample is found, the random number somehow has to be extracted from the contract execution, s.t. other validators can verify the counterexample
- some special failure operation may be needed that returns the that value
- as representation of a counterexample is still an open question, there are not specifics yet as to how this instruction is supposed to look like (or if it is needed at all)

@@

\paragraph{Assertion contracts}
\lstref{lst:contract_templ} shows a possible template for an assertion contract - the placeholder \texttt{<?>} marks the locations where the individual code is inserted. Lines 3-5 prepare the stack, s.t. the parameter and storage are the two top elements. The lines below the placeholder for the body clean up the stack and leave it according to calling convention, i.e., the program returns an empty list of internal operations and the unit storage.
\lstinputlisting[label=lst:contract_templ, language=Michelson, caption=Template code for assertion contracts in Michelson, numbers=left]{listings/assertion_template.tz}

@@

\paragraph{Evaluation}
Compiling to Michelson directly is beneficial due of several reasons - the pipeline stays streamlined and outputs the final contracts ready for origination. It does not introduce dependencies to external tools and their support for new protocol versions of Tezos. Besides, the new instructions only have to be added to Tezos' protocol code, whereas using external tools and languages would need to be patched as well in order to support them. \\
When it comes to verifying the compiler and integration testing of the pipeline, however, the low-level output causes more complexity than a high-level output. Due to the extend and restricted readability of non-trivial Michelson programs, their correctness is not always apparent in an instant. Tests can, on the one hand, compare the output with verified target contracts for a range of inputs, which requires a lot of Michelson code written by hand for each input. They, in turn, can also be prone to errors. On the other hand, the generated contracts can be tested as a black box using the test frameworks provided by Tezos, which use sandboxed networks to interact with contracts-under-test.

@@

\paragraph{Assertion contracts}
\lstinputlisting[label=lst:contract_templ_liq, language=Michelson, caption=Template code for assertion contracts in Liquidity, numbers=left]{listings/assertion_template.liq}

\paragraph{Manager contract}
\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Manager code in Liquidity for an entrypoint with a non-empty assertion, numbers=left]{listings/manager_loop.liq}

\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Manager code in Liquidity for an entrypoint with an empty assertion, numbers=left]{listings/manager_forward.liq}

\paragraph{Evaluation}
+ optimization (Liq compiler)
+ testing: easier to verify correctness, as it reads like an ocaml program
	+ compiler of hl language already tested -> only check if IR is correct
+ variable support -> variables in assertion easier to implement?

\section{Misc}
\todo{Aspects/questions I don't know yet if or where to put it}
\begin{itemize}
\item Tezos specific costs analysis of chosen orchestration strategy -> PROBLEM: high n might reach operation gas limit?
\item Representation of counterexamples; how to verify them?
\item Basic approach for proofs?
\end{itemize}

>>>>

@@

\subsection{Gas model}
In Tezos data is stored and transmitted as byte sequences --- in order to obtain a typed representation of their value for interpretation, a byte sequence is first deserialised into Micheline and subsequently parsed \cite{tezos_repo}. Conversely, before transmitting or storing something on the blockchain, the typed representation is first unparsed and then serialised. The gas consumption of an operation, e.g. a transaction or an origination, is thus composed of the following positions on top of a base cost \cite{morley_gasmodel}\cite{tezos_repo}:
\begin{enumerate}
\item \textbf{Reading costs} --- apply for reading contract code, storage or Big Maps from the internal database and, on top of the base cost, depend on the amount of bytes read
\item \textbf{Deserialization costs} --- depend on the Micheline expression and their size
\item \textbf{Parsing costs} --- are composed of three types of parsing costs:
	\begin{itemize}
	\item type parsing --- apply when a Micheline node is converted into a valid type and are computed by a recursive formula; parametrized types therefore generate more cost than plain types
	\item data parsing --- apply when a node is converted into the value of a known type and depend on the byte length for non-parametrized types, on the number of contained elements for container types and on the parameters for parametrized types
	\item code parsing --- computes the return type of the code by sequentially applying instructions and type checking if the result matches the expectation.The cost thus increases in the number of instructions, while some instructions incur additional costs besides their base cost, such as \texttt{IF} (both branched need to be checked and compared) or \texttt{CREATE\_CONTRACT} (called contracts needs to be type checked)
	\end{itemize}
\item \textbf{Type comparison costs} --- occur during parsing when types are checked for equality (e.g. when verifying that the branches of \texttt{IF} have the same type) or contract execution
\item \textbf{Interpreter costs} --- apply for each executed instruction and are specified for each. Expensive instruction are, for instance, \texttt{CONTRACT} with a high base cost and additional cost for reading, deserializing and type checking of the target contract code. 
\item \textbf{Unparsing costs} --- depend on the type of the data; parametrized types incur cost for unparsing its inner types as well
\item \textbf{Serialization costs} --- depend, similar to deserialization, on the Micheline expression and their size
\item \textbf{Writing costs} --- apply when writing to the internal database. Depend on the amount of bytes written, but with a higher base cost than reading. Additional cost is inflicted if the storage of the contract increased in size
\end{enumerate}
If a transaction emits internal operations, their gas consumption is computed in the same way and are added to the total gas consumption of the external operation.

@@

\subsubsection{High-level languages}
Michelson is a compilation target for various high-level languages that provide a more user-friendly and intuitive way of writing smart contracts. Additionally, some of them come with development environments and testing or verification tools. The following list comprises of the three most prominent languages that compile to Michelson:
\begin{itemize}
\item[\textbf{Liquidity}] \cite{liquidity} is a language with an OCaml-like syntax, which allows using local variables instead of stack manipulations. Its module system can be used to write reusable contract code or libraries. Besides an optimizing compiler, the project also includes a decompiler to compile Michelson programs to Liquidity.
\item[\textbf{SmartPy}] \cite{smartpy} is a language available through a Python library and lets developers write contracts and tests using Python syntax and structures (such as classes). Its developer suite includes i.a. a compiler, a simulation engine for testing contracts and an online editor.
\item [\textbf{Ligo}] \cite{ligo} offers multiple syntax flavours that are close to Pascal, OCaml or Reason. Besides compilation, the toolchain allows to invoke or evaluate the generated code locally. The Ligo OCaml packages can be integrated into other OCaml projects, s.t. the compiler can be invoked from within third-party applications.
\end{itemize}

@@

\subsection{random}
A vital instruction that cannot be expressed with Michelson's current instruction set (state of protocol 008 Edo \cite{tezos_docs}) is the generation of a random value of certain types. Since smart contracts need to be deterministic, s.t. the network can reach a consensus about the state of the ledger \cite{chatterjee_probabilistic_2019}, the prevalent workarounds for generating pseudo-random numbers cannot be used for assertion checking. Distributed assertion checking explicitly requires as many validators as possible to generate a unique value, whereas common schemes for randomness, like oracles or using block attributes as seeds \cite{chatterjee_probabilistic_2019}, will provide the same values for all validators.\\
\lstref{lst:rand_type} specifies the selection and typing rule of a new instruction \texttt{rand}, which consumes an offset and a positive range from the stack and pushes a randomly generated integer to the top. This instruction, of course, can also be provided for other data types, such as \texttt{nat}, \texttt{string} or \texttt{mutez}.
\lstset{upquote=true}
\begin{lstlisting}[caption=Selection and typing rules of the integer \texttt{rand} instruction, language=, numbers=none, label=lst:rand_type]
:: int : nat: 'A -> int : 'A
> RAND/ offset : range : S  => int : S
\end{lstlisting}

@@

\subsection{nth}
An exception for not adding higher level instructions may be the more fundamental operation \texttt{nth} for accessing list elements, as it is used (sometimes more than once, ref. \eqref{eq:sorted_v2}) in many use cases. Adding an own predefined primitive for this reduces the origination costs of the assertion contract, given that its equivalent in the current instruction set requires the compiler to generate a workaround using iteration, for instance a loop or list iterator. As a first approximation, appendix \ref{apx:nth} shows a potential implementation of \texttt{nth} in Liquidity (\lstref{lst:nth_liq}) using a loop, and how the respective Michelson code looks like after compiling the contract with the Liquidity compiler (\lstref{lst:nth_tz}). \\
In case the \texttt{nth} is added as a predefined primitive to Michelson for the type \texttt{'a list}, its selection and typing rules are specified as follows:
\begin{lstlisting}[caption=Selection and type rule of the list \texttt{nth} instruction, language=, label=lst:nth_type]
:: (list 'a) : nat : 'A -> option 'a : 'A
> NTH / l : index : S  =>  Some 'a: S
     iff index is within bounds
> NTH / l : index : S  =>  None 'a: S
     iff index is out of  bounds
\end{lstlisting}
As Michelson lists are represented with OCamls list type within the VM \cite{tezos_repo}, the computation for \texttt{nth} could look as shown in \lstref{lst:nth_impl}.
\begin{lstlisting}[caption=Simplified evaluation of \texttt{nth} in the Michelson interpreter, language=, label=lst:nth_impl]
match (instruction, stack) with
...
| (Nth ({elements = []; _}, (_, rest))) ->
  return ((None, rest)
| (Nth ({elements = es; length}, (index, rest))) ->
  let l_length = of_int length in
  if index <= l_length
    then return (Some (List.nth es index), rest)
    else return (None, rest)
\end{lstlisting}
\lstset{upquote=false}

@@

\paragraph{Assertion contracts}
\lstref{lst:contract_templ} shows a possible template for an assertion contract --- the placeholder \texttt{<?>} marks the locations where the individual code is inserted. Lines 3-5 prepare the stack, s.t. the parameter and storage are the two top elements. The lines below the placeholder for the body clean up the stack and leave it according to calling convention, i.e., the program returns an empty list of internal operations and the unit storage.
\lstinputlisting[label=lst:contract_templ, language=Michelson, caption=Template code for assertion contracts in Michelson, numbers=left]{listings/assertion_template.tz}

@@

\paragraph{Evaluation}
Compiling to Michelson directly is beneficial due of several reasons --- the pipeline stays streamlined and outputs the final contracts ready for origination. It does not introduce dependencies to external tools and their support for new protocol versions of Tezos. Besides, the new instructions only have to be added to Tezos' protocol code, whereas using external tools and languages would need to be patched as well in order to support them. \\
When it comes to verifying the compiler and integration testing of the pipeline, however, the low-level output causes more complexity than a high-level output. Due to the extend and restricted readability of non-trivial Michelson programs, their correctness is not always apparent in an instant. Tests can, on the one hand, compare the output with verified target contracts for a range of inputs, which requires a lot of Michelson code written by hand for each input. They, in turn, can also be prone to errors. On the other hand, the generated contracts can be tested as a black box using the test frameworks provided by Tezos, which use sandboxed networks to interact with contracts-under-test.

@@

\paragraph{Assertion contracts}
\lstref{lst:contract_templ_liq} shows the equivalent of the assertion contract template given in \lstref{lst:contract_templ}. Since the Liquidity compiler takes care of the handling of the stack, it basically consists only of a storage type declaration, a function signature and the ``empty'' return type.
\lstinputlisting[label=lst:contract_templ_liq, language=Michelson, caption=Template code for assertion contracts in Liquidity, numbers=left]{listings/assertion_template.liq}

\paragraph{Manager contract}
Liquidity contracts are a list of OCaml-like functions that take a parameter and a storage. Instead of declaring the contract parameter type globally and nesting the code for each entrypoint, each entrypoint thus is declared as a separate function with a local parameter type. During compilation, the contract parameter type will then be composed according to the order of functions. 
Due to Algorithm \ref{alg:compile_manager} being left recursive, the output will automatically retain the correct order of entrypoints. In fact, the only adaptions necessary are the output of the two functions \texttt{generate\_loop} and \texttt{generate\_forward}, which should not only generate the code body, but return a coherent function. The function \texttt{entrypointA} in \lstref{lst:manager_templ_liq} is an example of such a function for an empty assertion. The non-empty version of the manager code can be found in appendix \ref{apx:apx:manager_liq}. As already noted, extracting a pure template for the code exceeds the scope of this section.
\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Manager code in Liquidity for an entrypoint with an empty assertion, numbers=left]{listings/manager_forward.liq}

\paragraph{Evaluation}
This approach takes advantage of the usually more sophisticated high-level language compiler; they may provide options for optimizing generated code and are already thoroughly tested \cite{liq_repo}\cite{ligo_repo}. In contract to low-level code, the output of the pipeline compiler is easier to be verified, as it can be read like, for instance, an OCaml program. Providing target contracts for integration tests can be done much more efficiently and more reliable. \\
Besides the amenities of these compilers, this approach has a crucial disadvantage of introducing an external dependency. The compiler of the respective language has to be extended with the additional Michelson instructions as well, requiring a fork of the project. Furthermore one has to rely on continuous support of new protocol versions in the root project and integrate the updates into the own codebase. All things considered, the effort for extending both Michelson and the compiler of the IR, followed by a continuous effort of version management, may outweigh the benefits of this approach.

When returning to the orchestration approach of generating a monolithic contract, using Liquidity as an IR would allow to decompile the parent contract using the decompiler of its toolchain. Since the Liquidity compiler handles the composition of the contract parameter type and the code nesting, the assertion function simply have to be appended to the parent code, which significantly decreases the compilation complexity.

\section{Cost analysis}
Tezos specific costs analysis of chosen orchestration strategy -> PROBLEM: high n might reach operation gas limit?

\section{Open questions}

\subsection{Representing counterexamples}
\draft{}
- the random number somehow has to be extracted from the contract execution, s.t. other validators can verify the counterexample
- failwith exposes top element of the stack in error message -> can be used to "return" the random values? Question for protocol design how to process it further
- if this is not possible, Michelson may have to be extended with a special failure message that can be handled differently
- assertion execution -> random number is generated
- if counterexample is found, the random number somehow has to be extracted from the contract execution, s.t. other validators can verify the counterexample
- open question how the validators can check the counterexample -> send piece of code with random values as parameter?
- incentives \& penalties part of the block chain protocol

### 2021-04-09 assertions/doc/thesis-bernhardt/chapters/appendix.tex
IR done; add cost model; cost analysis

\appendix
\chapter{Assertion Grammar in EBNF}\label{apx:grammar}
The following two versions of the grammar implement the prefix and infix notations for the assertion syntax. 
\todo{Rename mutez}

@@

>>>>

\appendix
\chapter{Michelson contract used for cost analysis}
\lstinputlisting[basicstyle=\linespread{1.0}\fontfamily{lmr}\selectfont\small,
				 backgroundcolor=\color{cverbbg},
				 linewidth=14cm,
				 xleftmargin=0.5cm,
				 frame=lr,
				 framesep=8pt,
				 framerule=0pt,
				 captionpos=b,
				 numbers=none,
				 language=Michelson,
				 caption=Foobar.tz]{listings/foobar.tz} \todo{Caption}

\chapter{Assertion Grammar in EBNF}\label{apx:grammar}
The following two versions of the grammar implement the prefix and infix notations for the assertion syntax. 
\todo{Rename mutez}

@@

\chapter{Manager contract code in Liquidity}\label{apx:manager_liq}
\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Manager code in Liquidity for an entrypoint with a non-empty assertion, numbers=left]{listings/manager_loop.liq}
